A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-17 tex/body.tex
more

The mapping $\Val$ is computed as follows.
%
With each equivalence class $\ec{u}$, we associate a fresh variable $\Varec{u}$
of the same type as $u$
%not occurring in $\tEc$,
and let $\Val\ec{u} = {\Varec{u}}$. Therefore, there are initially no
constraints on the values for any equivalence class $\ec{u}$. The mapping $\Val$
is refined by applying the following unfolding rule exhaustively:\strut
\[
\hbox{\(
\inferrule{
  {\Varec{u}} \in \FV( \Val )
  \betweenantes
  \const{C}( t_1, \ldots, t_n ) \in \ec{u}
  \betweenantes
  \const{C} \in \Ctr
}{
  \Val := \Val [{\Varec{u}} \mapsto \MU {\Varec{u}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} )]
}
\)
%\rn{Unfold}
}
\]
$\FV( \Val )$ denotes the union of the free variables occurring in $\Val$'s range,
and $\Val[x \mapsto t]$ denotes the substitution of $t$ for $x$ in $\Val$'s range.
It is easy to see that the height of terms produced as a result of the unfolding
is bounded by the number of equivalence classes of $\Ec$,
and thus the construction of $\Val$ will terminate.

@@

First, we define the set of interpretations of the theory $\thD$,
which requires additional terminology concerning $\mu$-terms.
Given a $\mu$-term $t$ with subterm $u$,
%%% @ANDY: I'm taking the liberty of renaming "interpretation" to "expansion". Interpretation
%%% sounds semantic to me (although it is sometimes used for syntactic transformations).
the \emph{expansion of $u$ with respect to $t$} is the $\mu$-term $\interp{t}{u}^\emptyset$,
abbreviated to
$\interp{t}{u}$, as returned by the following recursive function:
%where $x \leadsto \MU x.\; \const{C}( \vec u ) \in t$ indicates that $\MU x.\;
%\const{C}( \vec u )$ is the subterm of $t$ that binds this occurrence of
%variable $x$:
\[\begin{array}{r@{}c@{}l}
\interp{t}{x}^B & {} \,=\, {} &
\textstyle\relax{\begin{cases}
    \MU x.\; \const{C}\smash{\bigl( \interp{t}{ \vec u}^{\smash{B \,\cup\, \{ x \}}}\bigr)} & \text{if this $x \notin B$ is
    bound by $\MU x.\; \const{C}( \vec u )$ in $t$}  \\[-\jot]
    x & \text{otherwise}
\end{cases}} ~ \\
\interp{t}{\MU x.\; \const{C}( \vec u )}^B & = &
\textstyle\begin{cases}
    \MU x.\; \const{C}\smash{\bigl( \interp{t}{ \vec u }^{\smash{B \,\cup\, \{ x \}}}\bigr)} & \text{if }x \not\in B \\[-\jot]
    x & \text{otherwise}
\end{cases}
\end{array}\]

The $\mu$-term $u$
is a \emph{self-similar subterm} of~$t$ if
$u$ is a strict subterm of $t$,
$t$ is of the form $\MU x.\; \const{C}( t_1, \ldots, t_n )$,
$u$ it is of the form $\MU y.\; \const{C}( u_1, \ldots, u_n )$,
and $\interp{t}{ t_k } \aequiv \interp{t}{ u_k }$ for all $k \in [n]$.
The term $t$ is \emph{normal} if it does not contain a self-similar subterm
and all of its strict subterms are also normal.
For example, $t = \MU x.\; \const{C}( \MU y.\; \const{C}( y ) )$ is abnormal
because $\MU y.\; \const{C}( y )$ is a self-similar subterm of $t$.
Their arguments have the same expansion with respect to
$t$:
$\interp{t}{\MU y.\; \const{C}( y )} =
\MU y.\; \const{C}\bigl( \interp{t}{y}^{\smash{\{y\}}} \bigr) =
\MU y.\; \const{C}( y )$
is $\alpha$-equivalent to
$\interp{t}{y} =
\MU y.\; \const{C}\bigl( \interp{t}{y}^{\smash{\{y\}}} \bigr) =
\MU y.\; \const{C}( y )$.
The term $t = \MU x.\; \const{C}( \MU y.\; \const{C}( x ) )$ is also abnormal,
since $\MU y.\; \const{C}( x )$ is a self-similar subterm of $t$,
noting that
%
\begin{align*}
\interp{t}{\MU y.\; \const{C}( x )}
& = \MU y.\; \const{C}\bigl( \interp{t}{x}^{\smash{\{y\}}} \bigr) \\
& = \MU y.\; \const{C}\bigl( \interp{t}{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) ) }^{\smash{\{y\}}} \bigr) \\
& = \MU y.\; \const{C}\bigl( \MU x.\; \const{C}( \interp{t}{\MU y.\; \const{C}( x )}^{\smash{\{x,y\}}} ) \bigr) \\
& = \MU y.\allowbreak\; \const{C}( \MU x.\allowbreak\; \const{C}( y ) )
\end{align*}
is $\alpha$-equivalent to $\interp{t}{x} = t$.
% for the renaming $\{ x \mapsto y, y \mapsto x \}$.

For any $\mu$-term $t$ of the form $\MU x.\; \const C( \vec u )$, we can
construct its \emph{normal form} $\nf{t}$
by replacing all of the self-similar subterms of $t$ with $x$
%and by replacing all of the abnormal subterms of $t$ with their corresponding normal forms.
and by recursively normalizing $t$'s subterms.
For example, $\nf{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) )} = \MU x.\; \const{C}( x )$.

>>>>

The mapping $\Val$ is computed as follows.
%
With each equivalence class $\ec{u}$, we associate a fresh variable $\Varec{u}$
of the same type as $u$
%not occurring in $\tEc$,
and set $\Val\ec{u} := {\Varec{u}}$. Therefore, there are initially no
constraints on the values for any equivalence class $\ec{u}$. The mapping $\Val$
is refined by applying the following unfolding rule exhaustively:\strut
\[
\hbox{\(
\inferrule{
  {\Varec{u}} \in \FV( \Val )
  \betweenantes
  \const{C}( t_1, \ldots, t_n ) \in \ec{u}
  \betweenantes
  \const{C} \in \Ctr
}{
  \Val := \Val [{\Varec{u}} \mapsto \MU {\Varec{u}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} )]
}
\)
%\rn{Unfold}
}
\]
$\FV( \Val )$ denotes the union of the free variables occurring in $\Val$'s range,
and $\Val[x \mapsto t]$ denotes the substitution of $t$ for $x$ in $\Val$'s range.
It is easy to see that the height of terms produced as a result of the unfolding
is bounded by the number of equivalence classes of $\Ec$,
and thus the construction of $\Val$ will terminate.

@@

First, we define the set of interpretations of the theory $\thD$,
which requires additional terminology concerning $\mu$-terms.
Given a $\mu$-term $t$ with subterm $u$,
%%% @ANDY: I'm taking the liberty of renaming "interpretation" to "expansion". Interpretation
%%% sounds semantic to me (although it is sometimes used for syntactic transformations).
the \emph{expansion of $u$ with respect to $t$} is the $\mu$-term $\interp{t}{u}^\emptyset$,
abbreviated to
$\interp{t}{u}$, as returned by the following recursive function:
%where $x \leadsto \MU x.\; \const{C}( \vec u ) \in t$ indicates that $\MU x.\;
%\const{C}( \vec u )$ is the subterm of $t$ that binds this occurrence of
%variable $x$:
\[\begin{array}{r@{}c@{}l}
\interp{t}{x}^B & {} \,=\, {} &
\textstyle\relax{\begin{cases}
    \MU x.\; \const{C}\smash{\bigl( \interp{t}{ \vec u}^{\smash{B \,\cup\, \{ x \}}}\bigr)} & \text{if this $x \notin B$ is
    bound by $\MU x.\; \const{C}( \vec u )$ in $t$}  \\[-\jot]
    x & \text{otherwise}
\end{cases}} ~ \\
\interp{t}{\MU x.\; \const{C}( \vec u )}^B & = &
\textstyle\begin{cases}
    \MU x.\; \const{C}\smash{\bigl( \interp{t}{ \vec u }^{\smash{B \,\cup\, \{ x \}}}\bigr)} & \text{if }x \not\in B \\[-\jot]
    x & \text{otherwise}
\end{cases}
\end{array}\]
%
\rem{TODO: add small examples}

The $\mu$-term $u$
is a \emph{self-similar subterm} of~$t$ if
$u$ is a proper subterm of $t$,
$t$ is of the form $\MU x.\; \const{C}( t_1, \ldots, t_n )$,
$u$ it is of the form $\MU y.\; \const{C}( u_1, \ldots, u_n )$,
and $\interp{t}{ t_k } \aequiv \interp{t}{ u_k }$ for all $k \in [n]$.
The term $t$ is \emph{normal} if it does not contain a self-similar subterm
and all of its proper subterms are also normal.
For example, $t = \MU x.\; \const{C}( \MU y.\; \const{C}( y ) )$ is abnormal
because $\MU y.\; \const{C}( y )$ is a self-similar subterm of $t$.
Their arguments have the same expansion with respect to~$t$:
$\interp{t}{\MU y.\; \const{C}( y )} =
\MU y.\; \const{C}\bigl( \interp{t}{y}^{\smash{\{y\}}} \bigr) =
\MU y.\; \const{C}( y )$
is $\alpha$-equivalent to
$\interp{t}{y} =
\MU y.\; \const{C}\bigl( \interp{t}{y}^{\smash{\{y\}}} \bigr) =
\MU y.\; \const{C}( y )$.
The term $t = \MU x.\; \const{C}( \MU y.\; \const{C}( x ) )$ is also abnormal,
since $\MU y.\; \const{C}( x )$ is a self-similar subterm of $t$,
noting that
%
\begin{align*}
\interp{t}{\MU y.\; \const{C}( x )}
& = \MU y.\; \const{C}\bigl( \interp{t}{x}^{\smash{\{y\}}} \bigr) \\
& = \MU y.\; \const{C}\bigl( \interp{t}{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) ) }^{\smash{\{y\}}} \bigr) \\
& = \MU y.\; \const{C}\bigl( \MU x.\; \const{C}( \interp{t}{\MU y.\; \const{C}( x )}^{\smash{\{x,y\}}} ) \bigr) \\
& = \MU y.\allowbreak\; \const{C}( \MU x.\allowbreak\; \const{C}( y ) )
\end{align*}
is $\alpha$-equivalent to $\interp{t}{x} = t$.
% for the renaming $\{ x \mapsto y, y \mapsto x \}$.

For any $\mu$-term $t$ of the form $\MU x.\; \const C( \vec u )$,
its \emph{normal form} $\nf{t}$ is obtained
by replacing all of the self-similar subterms of $t$ with $x$
%and by replacing all of the abnormal subterms of $t$ with their corresponding normal forms.
and by recursively normalizing the subterms.
For variables, $\nf{x} = x$.
Thus, $\nf{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) )} = \MU x.\; \const{C}( x )$.
